Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-06 _tex/paper2.tex
Built site for gh-pages

In its most general form, Thurstone's theory addresses pairwise
comparisons where a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the DP for any given stimulus
forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the mDP for Text B is positioned
further along the continuum than that of Text A \((T_{B} > T_{A})\),
suggesting that Text B exhibits higher quality. Additionally, the figure
highlights that Text B has a broader distribution compared to Text A,
which arises from its larger discriminal dispersion
\((\sigma_{B} > \sigma_{A})\).

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces \emph{the law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as \emph{the discriminal difference},
determines the observed dichotomous outcome. Moreover, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion rests on two key observations: the positive
difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

Thurstone noted that the general form of the theory, as delineated in
Section~\ref{sec-thurstone_theory}, gave rise to a trait scaling
problem. The model required estimating more ``unknown'' parameters than
the available pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To
address this issue and facilitate the practical implementation of the
theory, he developed five cases derived from this general form, each
case progressively incorporated additional simplifying assumptions into
the model.

@@

Despite relying on the most extensive set of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one notable distinction: whereas Case V
assumes a Normal distribution for the stimuli's discriminal processes,
the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

@@

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Therefore, it is reasonable to assume that the outcome's ability to
reflect the ``true'' differences between stimuli largely depends on the
validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
particularly the assumption of equal dispersions. For instance, when
researchers observe a sample of outcomes favoring Text B over Text A and
correctly assume equal dispersions between the texts, the BTL model
estimates a discriminal difference distribution that accurately
represents the ``true'' discriminal difference of the texts. This
scenario is illustrated in Figure~\ref{fig-dispersion}, where the
model's discriminal difference distribution aligns with the ``true''
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
difference ensures reliable estimates for the texts' discriminal
processes {(citation needed?)}.

Nevertheless, model misspecification, in the form of an erroneous
assumption of equal dispersions between stimuli, can give rise to
significant statistical and measurement issues. For instance, the model
may overestimate the degree to which the outcome accurately reflects the
``true'' discriminal differences between stimuli. This overestimation
can result in researchers drawing spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
illustrates this issue when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).

Additionally, if researchers recognize that misfit statistics highlight
these critical differences in dispersions, the conventional CJ practice
of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information. Such exclusions can
introduce bias into trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases are often unpredictable, as they depend on
which stimuli are excluded from the analysis.

@@

@@

The correlation, represented by the symbol \(\rho\), measures how much a
judge's perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} presents a similar thought experiment as in
Section~\ref{sec-theory-issue1a} to illustrate this idea. The
illustration now assumes that the discriminal dispersions for both texts
remain constant. The figure shows that as the correlation between the
texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing affects the area under the curve where
the discriminal difference distinctly favors Text B over Text A, denoted
as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
the figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor Text
B over Text A (shaded gray area) compared to scenarios where the texts
are highly correlated.

Off course, in experimental practice, researchers approach this process
in reverse. They begin by observing the sample of outcomes favoring Text
B over Text A and then use the BTL model to estimate the discriminal
difference and the discriminal processes of the stimuli. Given that the
BTL model assumes independent discriminal processes across comparisons,
if this assumption holds, then the model estimates a discriminal
difference distribution that accurately reflects the ``true''
discriminal difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the estimation accuracy of the discriminal difference ensures
reliable estimates for the discriminal processes of the texts {(citation
needed?)}.

@@

In any case, similar to Section~\ref{sec-theory-issue1a}, model
misspecification due to an erroneous assumption of zero correlation
between stimuli can lead to significant statistical and measurement
issues. For instance, the model may over- or underestimate how
accurately the outcome reflects the ``true'' discriminal differences
between stimuli. Such inaccuracies can result in spurious inferences
about these differences and, by extension, about the stimuli's
discriminal processes. This scenario is illustrated in
Figure~\ref{fig-correlation}, where the model's discriminal difference
distribution aligns with the thick continuous line for \(\rho=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\rho \neq 0\).

This misspecification may arise from neglecting additional relevant
traits, excluding judges based on misfit statistics, or ignoring
hierarchical (grouping) structures. Neglecting relevant traits, such as
judges' biases, can cause dimensional mismatches in the BTL model,
artificially inflating the trait's reliability
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introducing bias into the
trait's estimates \citep{Ackerman_1989}. Excluding judges based on
misfit statistics risks discarding valuable information, which may
further bias the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
hierarchical structures may reduce the precision of model parameter
estimates, potentially amplifying the overestimation of the trait's
reliability \citep[pp.~482]{Hoyle_et_al_2023}.

@@

This section presents a theoretical model for CJ that extends
Thurstone's theory. It uses causal analysis
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014, Hernan_et_al_2020}
and, in particular, directed acyclic graphs (DAGs)
\citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
principles of CJ theory. The model also incorporates several practical
factors that influence judges in CJ experiments, such as the selection
of judges, stimuli, and the number of comparisons. In addition, the
study uses Bayesian statistical methods \citep{McElreath_2020} to
translate these theoretical and practical elements into a statistical
model that facilitates the analysis of pairwise comparison data.

@@

\paragraph{Figure}

\paragraph{3}

@@

\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}

@@

\caption{\label{fig-CJ_TM_A1}}

@@

\paragraph{4}

@@

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}

@@

\caption{\label{fig-CJ_TM_A2}}

@@

\paragraph{5}

@@

\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}

@@

\caption{\label{fig-CJ_TM_A3}}

@@

\paragraph{6}

@@

\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A4.png}

@@

\caption{\label{fig-CJ_TM_A4}}

\paragraph{Figure}

\paragraph{7}

\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B1.png}

\caption{\label{fig-CJ_TM_B1}}

\paragraph{8}

\includegraphics[width=0.62\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B2.png}

\caption{\label{fig-CJ_TM_B2}}

\paragraph{9}

\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B3.png}

\caption{\label{fig-CJ_TM_B3}}

\paragraph{10}

\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B4.png}

\caption{\label{fig-CJ_TM_B4}}

\paragraph{11}

\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B5.png}

\caption{\label{fig-CJ_TM_B5}}

\paragraph{12}

\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B6.png}

\caption{\label{fig-CJ_TM_B6}}

Considering the sampling and comparison mechanisms

\paragraph{Figure}

\paragraph{13}

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B7.png}

\caption{\label{fig-CJ_TM_B7}}

\paragraph{14}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B8.png}

\caption{\label{fig-CJ_TM_B8}}

\paragraph{15}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B9.png}

\caption{\label{fig-CJ_TM_B9}}

>>>>

In its most general form, Thurstone's theory addresses pairwise
comparisons where a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the modal discriminal process for Text
B is positioned further along the continuum than that of Text A
\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as the \emph{discriminal difference},
determines the observed dichotomous outcome. Moreover, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion is supported by two key observations: the positive
difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

Thurstone noted from the outset that the general form of the theory, as
outlined in Section~\ref{sec-thurstone_theory}, gave rise to a problem
of trait scaling. The model required estimating more ``unknown''
parameters than the available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
the practical implementation of the theory, he developed five cases
derived from this general form, each case progressively incorporated
additional simplifying assumptions into the model.

@@

Notably, despite relying on the most extensive set of simplifying
assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one notable distinction: whereas Case V
assumes a Normal distribution for the stimuli's discriminal processes,
the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

@@

@@

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Therefore, the outcome's ability to reflect the ``true'' differences
between stimuli largely depends on the validity of the model's
assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
assumption of equal dispersions. For instance, when researchers observe
a sample of outcomes favoring Text B over Text A and correctly assume
equal dispersions between the texts, the BTL model estimates a
discriminal difference distribution that accurately represents the
``true'' discriminal difference of the texts. This scenario is
illustrated in Figure~\ref{fig-dispersion}, where the model's
discriminal difference distribution aligns with the ``true''
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
difference ensures reliable estimates for the texts' discriminal
processes {(citation needed?)}.

Thus, model misspecification, in the form of an erroneous assumption of
equal dispersions between stimuli, can give rise to significant
statistical and measurement issues. For instance, the model may
overestimate the degree to which the outcome accurately reflects the
``true'' discriminal differences between stimuli. This overestimation
can result in researchers drawing spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this issue when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
Additionally, if researchers recognize that misfit statistics highlight
these critical differences in dispersions, the conventional CJ practice
of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information. Such exclusions can
introduce bias into trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases are often unpredictable, as they depend on
which stimuli are excluded from the analysis.

@@

The correlation, represented by the symbol \(\rho\), measures how much a
judge's perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} presents a similar thought experiment as in
Section~\ref{sec-theory-issue1a} to illustrate this idea. The
illustration now assumes that the discriminal dispersions for both texts
remain constant. The figure reveals that as the correlation between the
texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing affects the area under the curve where
the discriminal difference distinctly favors Text B over Text A, denoted
as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
the figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor Text
B over Text A (shaded gray area) compared to scenarios where the texts
are highly correlated.

Off course, in experimental practice, researchers approach this process
in reverse. They begin by observing the sample of outcomes favoring Text
B over Text A and then use the BTL model to estimate the discriminal
difference and the discriminal processes of the stimuli. Given that the
BTL model assumes independent discriminal processes across comparisons,
if this assumption holds, then the model estimates a discriminal
difference distribution that accurately reflects the ``true''
discriminal difference of the texts. This scenario is also illustrated
in Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the estimation accuracy of the discriminal difference ensures
reliable estimates for the discriminal processes of the texts {(citation
needed?)}.

@@

In any case, similar to Section~\ref{sec-theory-issue1a}, model
misspecification due to an erroneous assumption of zero correlation
between stimuli can lead to significant statistical and measurement
issues. For instance, the model may over- or underestimate how
accurately the outcome reflects the ``true'' discriminal differences
between stimuli. Such inaccuracies can result in spurious inferences
about these differences and, by extension, about the stimuli's
discriminal processes. This scenario is also illustrated by
Figure~\ref{fig-correlation}, when the model's discriminal difference
distribution aligns with the thick continuous line for \(\rho=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\rho \neq 0\).

The misspecification may arise from neglecting additional relevant
traits, excluding judges based on misfit statistics, or ignoring
hierarchical (grouping) structures. Neglecting relevant traits, such as
judges' biases, can cause dimensional mismatches in the BTL model,
artificially inflating the trait's reliability
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introducing bias into the
trait's estimates \citep{Ackerman_1989}. Excluding judges based on
misfit statistics risks discarding valuable information, which may
further bias the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
hierarchical structures may reduce the precision of model parameter
estimates, potentially amplifying the overestimation of the trait's
reliability \citep[pp.~482]{Hoyle_et_al_2023}.

@@

This section presents a theoretical model for CJ that extends
Thurstone's theory. It uses causal analysis
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014} and, in
particular, directed acyclic graphs (DAGs)
\citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
principles of CJ theory. The model also incorporates several practical
factors that influence judges in CJ experiments, such as the selection
of judges, stimuli, and comparisons. In addition, the study uses
Bayesian statistical methods \citep{McElreath_2020} to translate these
theoretical and practical elements into a statistical model that
facilitates the analysis of pairwise comparison data.

@@

@@

\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}

@@

\caption{\label{fig-CJ_TM_04}}

@@

@@

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}

@@

\caption{\label{fig-CJ_TM_10}}

@@

Considering the sampling mechanism

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}

@@

\caption{\label{fig-CJ_TM_12}}

@@

Considering comparison mechanisms

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}

@@

\caption{\label{fig-CJ_TM_14}}
